---
title: |
  | DSC 152: 
  | Applied Statistical Data Analysis and Inference
author: | 
  | Lecture \#17 Part 2
  | Introduction to 
  | Time Series Analyses
output: 
  beamer_presentation:
    theme: "Warsaw"
    fonttheme: "serif"

header-includes:
- \setbeamertemplate{footline}[frame number]{}
- \setbeamertemplate{navigation symbols}{}
- \setbeamersize{text margin left=0.5cm,text margin right=0.5cm}
- \definecolor{codegray}{HTML}{f4f4f4}
- \let\textttOrig\texttt
- \renewcommand{\texttt}[1]{\textttOrig{\colorbox{codegray}{#1}}}
---
    
```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_chunk$set(dev = 'pdf')
  def.chunk.hook  <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
  })
```


## Moving Average Models {.t}
In a Moving Average model, each value is a function of a *weighted average of past values.*

::: {.t}
### Examples
\vspace{-2mm}

 - A period of high demand at a store results in the store increasing their inventory; if demand drops for long enough, they will decrease their inventory
 
\vspace{3mm}

 - Prolonged weather patterns may influence foot traffic at retail locations, impacting their sales. 
:::

Moving Average models are referenced as ~~MA(p)~~ MA(q) where ~~p~~ q is the number of past values that influence the current value.

\textit{(apologies for the typo, I originally had this as MA(p) in the slides)}

## Moving Average Models {.t}
\framesubtitle{What might an MA(q) process look like?}

Let's simulate some values according to an MA(5) process and plot it:
```{r, eval=FALSE}
y <- arima.sim(model = 
                 list(ma = c(0.5, 0.4, 0.3, 0.2, 0.1)), 
               n = 200)
t <- 1:200
df <- data.frame(y=y, t=t)
ggplot(df, aes(x=t, y=y)) + 
  geom_line()
```

## Moving Average Models {.t}
\framesubtitle{What might an MA(q) process look like?}
```{r, echo=FALSE, message=FALSE}
library(ggplot2)
y <- arima.sim(model = list(ma = c(0.5, 0.4, 0.3, 0.2, 0.1)), n = 200)
t <- 1:200
df <- data.frame(y=y, t=t)
ggplot(df, aes(x=t, y=y)) + 
  geom_line()
```

## Moving Average Models {.t}
\framesubtitle{What might an MA(q) process look like?}
It's hard to discern from that plot! But what about an ACF plot?
```{r, out.width="85%"}
acf(y)
```

## Lucas 2013 paper


## Your Turn \#2 {.t}

 - Load in the data from the `casino_poker_data.csv` file on the course website
 
\vspace{4mm}

 - Make a timeplot of the `R2_COIN` variable
 
\vspace{4mm}

 - Make an ACF plot of the `R2_COIN` variable and briefly comment on what you observe.