---
title: |
  | DSC 152: 
  | Applied Statistical Data Analysis and Inference
author: | 
  | Lecture \#19
  | Time Series Regression
  | Power Estimation
date: |
  | Tuesday, June 2
  | Spring Quarter 2026
  | Peter Chi  
output: 
  beamer_presentation:
    theme: "Warsaw"
    fonttheme: "serif"

header-includes:
- \setbeamertemplate{footline}[frame number]{}
- \setbeamertemplate{navigation symbols}{}
- \setbeamersize{text margin left=0.5cm,text margin right=0.5cm}
- \definecolor{codegray}{HTML}{f4f4f4}
- \let\textttOrig\texttt
- \renewcommand{\texttt}[1]{\textttOrig{\colorbox{codegray}{#1}}}
---
    
```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  knitr::opts_chunk$set(dev = 'pdf')
  def.chunk.hook  <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
  })
```


## Recall yet again...
\framesubtitle{Conditions for validity of inference in linear regression}


 - The relationship between $X$ and $Y$, if there is one, is actually \underline{\textbf{L}inear}
   - e.g. not quadratic, exponential, etc.

\vspace{2mm}

 - \textbf{I}ndependence of observations

\vspace{2mm}

 - \textbf{N}ormality of $\epsilon_i$
   - Note that this can also be achieved, due to the central limit theorem, with a large sample size even if $\epsilon_i$ does not follow a normal distribution

\vspace{2mm}

 - \textbf{E}qual variance across all values of $X$
   - Also known as homoskedasticity
   
## How much does it actually matter? {.t}
What happens if the data actually are time-dependent, but we ignore that in our analysis? 

### Recall the poker example...

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="75%"}
library(dplyr)
library(readr)
library(ggplot2)
export_all <- read_csv("ReportExport.csv")
won_all <- gsub("\\$", "", export_all$`My C Won`)
won_all <- gsub("\\(", "-", won_all)
won_all <- gsub("\\)", "", won_all)
won_all <- as.numeric(won_all)

df_all <- data.frame(won_all = won_all,
                     hand = 1:length(won_all))

df_all <- df_all %>% mutate(won_cum = cumsum(won_all))

ggplot(data=df_all, aes(x=hand, y=won_cum)) + 
  geom_line() + 
  labs(title="Running total of money won starting from May 7, 2020", y = "Cumulative amount won") + 
  scale_y_continuous(labels = scales::label_dollar()) +
  theme(text = element_text(size=20))
```



## Simulation under an AR(p) model {.t}

::: {.t}
### Recall: an AR(2) model
\vspace{-6mm}
$$
y_t = c + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
$$
:::

Now suppose the data generating process is this AR(2) model with:

 - $c=2$
 - $\phi_1 = 0.65$
 - $\phi_2 = 0.25$
 - $\epsilon_t \sim N(0, 1)$

::: {.t}
### which would give:
\vspace{-6mm}
$$
y_t = 2 + 0.65 y_{t-1} + 0.25 y_{t-2} + \epsilon_t
$$
:::


Recall that $c$ is the drift parameter...



## Simulation under an AR(p) model {.t}
```{r, echo=FALSE, cache=TRUE}
library(ggplot2)
set.seed(3)
eps <- rnorm(102, 0, 1)

# We need two burn-in values
y <- 2 + eps[1]
y[2] <- 2 + 0.65*y + eps[2]
for(i in 3:102){
  y[i] <- 2 + 0.65*y[i-1] + 0.25*y[i-2] + eps[i]
}

time <- 1:100
y <- y[3:102] # discard the burn-in
df <- data.frame(time=time, y = y)

ggplot(df, aes(x=time, y=y)) + geom_line()
```

## Simulation under an AR(p) model {.t}
```{r, eval=FALSE}
library(ggplot2)
set.seed(3)
eps <- rnorm(102, 0, 1)

# We need two burn-in values
y <- 2 + eps[1]
y[2] <- 2 + 0.65*y + eps[2]
for(i in 3:102){
  y[i] <- 2 + 0.65*y[i-1] + 0.25*y[i-2] + eps[i]
}

time <- 1:100
y <- y[3:102] # discard the burn-in
df <- data.frame(time=time, y = y)

ggplot(df, aes(x=time, y=y)) + geom_line()
```

## Simulation under an AR(p) model {.t}
\label{diffs}
What would a t-test give on the differences for this particular simulation?
```{r, results='hide'}
diffs <- NULL
for(i in 1:99){
  diffs[i] <- y[i+1] - y[i]
}

t.test(diffs)
```

Wait why on the differences?


## Simulation under an AR(p) model {.t}
For this particular simulation:

```{r}
t.test(diffs)
```


## Simulation under an AR(p) model {.t}
```{r, message=FALSE}
library(forecast)
library(lmtest)

model2 <- Arima(y, order=c(2,0,0), xreg=time)
coeftest(model2)
```


## Simulation under an AR(p) model {.t}

::: {.t}
### What did we observe?
\vspace{-2mm}

In this particular case, we fail to reject $H_0$ with the t-Test, but correctly reject $H_0$ when we run the AR(2) model.

\vspace{5mm}

What would happen on average?
:::


## Simulation under an AR(p) model {.t}
\framesubtitle{Your Turn \#1}

::: {.t}
### First let's write two functions:
\vspace{-2mm}

 - One to take in inputs of:
   - $n$
   - $\phi_1$
   - $\phi_2$
   - and then output a dataframe of `y` and `time`. You can hardcode `c=2` and `sd=1`.

\vspace{1mm}   

 - One to calculate `diffs` as in Slide \ref{diffs}
 
:::

::: {.t}
### Then, use the function...
\vspace{-2mm}
to run a simulation study of 1000 reps with $n=100, \phi_1=0.65, \phi_2=0.25$ and compare power with:

 - the `t.test` (ignoring the AR(2) structure of the data)
 - `Arima` (properly accounting for the AR(2) structure of the data)

:::

Finally, repeat with $\phi_1=-0.65, \phi_2=-0.25$.


## Now, how about in Time Series *Regression*? {.t}

::: {.t}
### Recall the Jalen Brunson example from last time... 
\vspace{-2mm}

 - The covariate of interest was the opponent's defensive rating
 - In the simulated example, we generated data under an MA(1) model
 - We observed that in the analysis with `lm`, we failed to reject $H_0$, whereas in the analysis with `Arima`, we correctly rejected $H_0$. 

:::

But that was just one iteration. What happens on average?

Let's stick with an AR(2) model as in the previous example from today, and simulate without context...

## Now, how about in Time Series *Regression*? {.t}

::: {.t}
### Suppose the true model is:
\vspace{-6mm}
$$
y_t = \beta_0 + \beta_1 x_t + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
$$
:::

where:

 - $\beta_0 = 0$
 - $\beta_1 = 2$
 - $\phi_1 = 0.65$
 - $\phi_2 = 0.25$
 - $\epsilon_t \sim N(0, 3)$
 
Let's do a simulated power comparison!


## Now, how about in Time Series *Regression*? {.t}
\framesubtitle{Your Turn \#2}
First, together we will write a new function that will be a modification of the `gen_AR2` function from Your Turn \#1. 

 - $\beta_0, \beta_1$ and the distribution of $\epsilon_t$ can be hardcoded. 
 - Let the $x$ values come from a Unif(0, 10) distribution (which can also be hardcoded)
 - You'll need two burn-in values in a similar manner to that of Your Turn \#1
 

## Now, how about in Time Series *Regression*? {.t}
\framesubtitle{Your Turn \#2}
Next, use that function to write a simulation comparing:

 - the statistical power with the `lm` function (ignoring the time dependency structure)
 
\vspace{5mm}

 - the statistical power with `Arima` (accounting for the time dependency structure)

## Recap and Looking Ahead {.t}

::: {.t}
### Recap
\vspace{-2mm}
Failing to account for time dependency in your data can lead to incorrect inference. In every example here, we showed a loss of power, but other things could happen.

:::

\vspace{4mm}

::: {.t}
### Looking Ahead
\vspace{-2mm}

 - Thursday's class: Final Exam Review
 - Saturday: Final Exam from 3-6pm. EVERYONE will be in SOLIS 107
 
:::


\vspace{4mm}


::: {.t}
### Today's Daily Check
\vspace{-2mm}
The two Your Turns
:::


## Quiz 3 Review